Work Fragmentation in Developer Interaction Data

### 2016-02-10 paper_2.tex
RQ3

The impact of work fragmentation could be more noticeable in the immediate minutes before and after an interruption occurs. On one hand, after an interruption the programmer carry out activities meant to recover the lost mental state, like reading the code, debugging, reading notes and cues, and more resumption strategies \cite{PR11}. Depending on the interruption, the recovery could last about 15 minutes, according to the reports from field studies \cite{IH07, SBV98}. In the data from UDC, the sessions with duration of at least seven hours (commonly the working time of a software developer) have 11 interruptions in average; if we consider a recovery time of 15 minutes, the total time required to recover from the interruptions can be of about two hours, which is about one quarter of the whole session.

%\RR{expand this paragraph, since a reviewer asks explicitely for that}
On the other hand, before the interruption there is a preparation phase when the interruption is imminent or expected, e.g. lunch time, a meeting and other scheduled events. In this phase the programmer might leave notes or comments in the code to recover the context after the interruption \cite{PD10}. Also, when the interruption happened because of a problem found by the programmer usually he tries to solve it by reading the code, using the debugger or reading comments in the control version tool; after this resources are depleted, the next option is to ask to teammates or other external resources \cite{LVD06}, which generates an interruption of work. The activities prior to the interruption are commonly associated with selection and navigation tasks, like reading the code and navigating to another classes or files to try to find an answer to the problem.

In this research question we expect to see the effects on productivity around an interruption due to the activities previously described.

@@

To summarize the development sessions, we have resized each time series to a single $size$, using local means. We have used $size = 10$, that is, dividing each session in 10 chunks of equal time, since because it yielded better global visualization (better uniformity) of the user interaction along a session. For each chunk, we choose the median value of edits and selections for each group, and compose one summary time series for each group; these are the time series displayed in Figure \ref{fig:bins_int}, for Mylyn, and Figure \ref{fig:bins_int_udc}, for UDC.

@@

\item \textit{On the right,} the trend of the time series increases steadily. We hypothesize that the programmer is immersing again into the programming task, increasing progressively the activity as represented by the number of edits and selections. In Mylyn it reaches the median activity (in average) 12 to 18 minutes after the interruption, while in UDC it is around 6 to 12 minutes. It then rises further than the session median, which is not surprising, as we expect higher than median activity further away from interruptions.

%We called \textit{recuperation time} the first intersect point between edition and navigation. In this sense, the  average recuperation time of an interruption occurs in the minute 12. 
\item \textit{On the left,} we observe that the number of events near the interruption also goes down well below the median value. This might be because of two reasons: the programmer could have found a problem while coding, who at first will try to solve it by reading the code, switching out from the IDE, navigating the call stack and debugging, before going to ask another teammate according to LaToza et al. \cite{LVD06}; this set of actions end up with an interruption and reduce the observed activity within the IDE. Also, when the interruption is imminent or expected, the programmer make use of different suspension strategies like writing physical notes, making a mental note or leaving a reminder cue on the code or window, as mentioned by Parnin and DeLine \cite{PD10}. These activities reduce the activity before the gap and they are seen in the interaction data mostly as selection events. The latest can be better visualized in Figure \ref{fig:all_ints} with the UDC data as a significant decrease of edit events, being below the selection events approximately 10 minutes before the interruption.
%\item \textit{On the left,} we observe that the number of events near the interruption also goes down well below the median value. This is because work fragmentation is not only due to interruptions: programmers may for instance switch away from the IDE to look up information on the web or request it from other programmers. They are likely to do so because they are stuck, which would manifest itself by lower activity before the activity gap. We investigate this further below.

@@

We did so by computing Cohen's $d$ on the quantity of edits before and after the interruption. To obtain a significant effect, we need the presence of activity both before and after. Not all the interruptions meet this criteria however: some are located close to the start or the end of a session, or too close to another interruption.  In total, 53\% of the interruptions had 30 minutes before and after the interruption and were selected for the analysis.  The remaining 47\% of the interruptions can not be used on this analysis due to the lack of time around them. %We used the Algorithm \ref{alg:local_effect} for the classification, and the 
Table \ref{tbl:local_effect} shows the applied thresholds and the results, accompanied with a typical example of an interruption in each category. This local analysis shows that there are indeed three well-defined groups of interruptions, with the two largest of them having clear effects on the activity in the session.

@@

In contrast, when facing the effects of a negative interruption the metrics take longer to reach the median. On the contrary to facing a positive interruption, the editions per minute metric is below the selections per minute, and the former takes longer than the latter to reach the average value. Also, the editions per minute metric is above the average for the 30 minutes before the interruption.

In general, the effect of positive and negative interruptions is the inverse. The editions per minute describes better the effect and the selections do not show a major change. There is not a clear pattern during a neutral interruption.

@@

>>>>

The impact of work fragmentation could be more noticeable in the immediate minutes before and after an interruption occurs. On the one hand, after an interruption the programmer may carry out activities meant to recover the lost mental state, such as reading the code, debugging, reading notes and cues, among other resumption strategies \cite{PR11}. Depending on the interruption, the recovery could last about 15 minutes, according to the reports from field studies \cite{IH07, SBV98}. In the data from UDC, the sessions with duration of at least seven hours (commonly the working time of a software developer) have 11 interruptions in average; if we consider a recovery time of 15 minutes, the total time required to recover from the interruptions can be of about two hours, which is about one quarter of the whole session.

%\RR{expand this paragraph, since a reviewer asks explicitely for that}
On the other hand, before the interruption there is a preparation phase when the interruption is imminent or expected, e.g. lunch time, a meeting and other scheduled events. In this phase the programmer might leave notes or comments in the code to recover the context after the interruption \cite{PD10}. Also, when the interruption happened because of a problem found by the programmer, the programmer may try to solve it by reading the code, or using the debugger;%or reading comments in the control version tool
after these resources are depleted, the next option is to ask to teammates or other external resources \cite{LVD06}, which generates an interruption in the work. The activities prior to the interruption are commonly associated with selection and navigation tasks, such as reading the code and navigating to another classes or files to try to find an answer to the problem.

In this research question we expect to see the effects on productivity around an interruption due to the activities previously described, and verify this.

@@

To summarize the development sessions, we have resized each time series to a single $size$, using local means. We have used $size = 10$, that is, dividing each session in 10 chunks of equal time, since this yielded better global visualization (better uniformity) of the user interaction along a session. For each chunk, we choose the median value of edits and selections for each group, and compose one summary time series for each group; these are the time series displayed in Figure \ref{fig:bins_int}, for Mylyn, and Figure \ref{fig:bins_int_udc}, for UDC.

@@

\item \textit{On the right,} the trend of the time series increases steadily. We hypothesize that the programmer is immersing again into the programming task, increasing progressively the activity as represented by the number of edits and selections. In Mylyn it reaches the median activity (in average) 12 to 18 minutes after the interruption, while in UDC this takes around 6 to 12 minutes. It then rises further than the session median, which is not surprising, as we expect higher than median activity further away from interruptions.

%We called \textit{recuperation time} the first intersect point between edition and navigation. In this sense, the  average recuperation time of an interruption occurs in the minute 12. 
\item \textit{On the left,} we observe that the number of events near the interruption also goes down well below the median value. This was more surprising at first to us. However we hypothesize that this might be because of two reasons: the programmer could have found a problem while coding, and at first would try to solve it by reading the code, switching out from the IDE, navigating the call stack and debugging, before going to ask another teammate (as documented e.g. by LaToza et al. \cite{LVD06}); this set of actions end up with an interruption and reduce the observed activity within the IDE. In the case when the interruption is imminent or expected, another possibility is that the programmer may make use of different suspension strategies such as writing physical notes, making a mental note or leaving a reminder cue on the code or window, as documented by Parnin and DeLine \cite{PD10}. These activities reduce the activity before the gap and they are seen in the interaction data mostly as selection events. The latest can be better visualized in Figure \ref{fig:all_ints}, in which the UDC data has a significant decrease of edit events, being below the selection events approximately 10 minutes before the interruption.
%\item \textit{On the left,} we observe that the number of events near the interruption also goes down well below the median value. This is because work fragmentation is not only due to interruptions: programmers may for instance switch away from the IDE to look up information on the web or request it from other programmers. They are likely to do so because they are stuck, which would manifest itself by lower activity before the activity gap. We investigate this further below.

@@

We did so by computing Cohen's $d$ on the quantity of edits before and after the interruption. To obtain a significant effect, we need the presence of activity both before and after the interruption. Not all the interruptions meet this criteria however: some are located close to the start or the end of a session, or too close to another interruption (in this, the next interruption would ``disturb'' the observation).  In total, 53\% of the interruptions had 30 minutes before and after the interruption and were selected for the analysis.  The remaining 47\% of the interruptions can not be used on this analysis due to the lack of time around them. %We used the Algorithm \ref{alg:local_effect} for the classification, and the 
Table \ref{tbl:local_effect} shows the applied thresholds and the results, accompanied with a typical example of an interruption in each category. This local analysis shows that there are indeed three well-defined groups of interruptions, with the two largest of them having clear effects on the activity in the session.

@@

In contrast, when facing the effects of a negative interruption the metrics take longer to reach the median. Contrary to when facing a positive interruption, the editions per minute metric is below the selections per minute, and the former takes longer than the latter to reach the average value. Also, the editions per minute metric is above the average for the 30 minutes before the interruption.

In general, the effect of positive and negative interruptions is seems to be reversed. The editions per minute describes better the effect and the selections do not show a major change. There is not a clear pattern during a neutral interruption.

@@

These observations give preliminary support to our hypotheses on the distinct types of interruptions that we evoked above. In RQ5, we delve deeper in this question by performing an exploratory analysis of the finer-grained types of activity found in each kind of session.
